---
title: "CTCF borders versus histone modifications and features - mESC cells"
author: "Tom van Schaik"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document:
    theme: journal #cerulean
    highlight: monochrome
    toc: true
    toc_float: true
    code_folding: show
  editor_options:
    chunk_output_type: console
---


### Project description

Effect of cohesin looping on genome architecture, from the perspective of the 
nuclear lamina.


### Introduction

Look whether CTCF borders are distinct from non-CTCF borders in mESC cells.


### Method

Deeptools.


### Set-up

Load the libraries and set the parameters.

```{r set-up, warning = F, message = F}

# Load dependencies
library(tidyverse)
library(GenomicRanges)
library(rtracklayer)
library(ggplot2)
#library(matrixStats)
#library(mgcv)
library(caTools)


# Prepare output 
output.dir <- "ts210202_CTCF_borders_versus_histone_modifications_mESC"
dir.create(output.dir, showWarnings = FALSE)

# Load input
input.dir <- "ts210201_CTCF_enrichment_at_LAD_borders"
metadata.borders <- readRDS(file.path(input.dir, "metadata.rds"))

LAD.borders <- readRDS(file.path(input.dir, "LAD_borders.rds"))
LAD.borders.mesc <- LAD.borders[["mESC"]]

# LAD.borders <- readRDS(file.path(input.dir, "LAD_borders_pA.rds"))
# LAD.borders.mesc <- LAD.borders[["mESC_pA_PT"]]

```

```{r knits setup}
library(knitr)
opts_chunk$set(fig.width = 10, fig.height = 4, cache = T,
               dev=c('png', 'pdf'), fig.path = file.path(output.dir, "figures/")) 
pdf.options(useDingbats = FALSE)
```

```{r functions}



```


### 1. Prepare LAD borders

Prepare the LAD borders for deeptools.

```{r prepare lad borders}

# Write mESC LAD borders with / without CTCF
LAD.borders.mesc.ctcf <- LAD.borders.mesc[LAD.borders.mesc$CTCF == "CTCF"]
LAD.borders.mesc.nonctcf <- LAD.borders.mesc[LAD.borders.mesc$CTCF != "CTCF"]

# Prepare file names and write bed files
LAD.borders.mesc.ctcf.file <- file.path(output.dir, "LAD_borders_mESC_CTCF.bed")
LAD.borders.mesc.nonctcf.file <- file.path(output.dir, "LAD_borders_mESC_nonCTCF.bed")

export.bed(LAD.borders.mesc.ctcf, LAD.borders.mesc.ctcf.file)
export.bed(LAD.borders.mesc.nonctcf, LAD.borders.mesc.nonctcf.file)

# Prepare list of files
LAD.borders.list <- c(LAD.borders.mesc.nonctcf.file, LAD.borders.mesc.ctcf.file)



# Also, do this for inwards / outwards borders
LAD.borders.mesc.outwards <- LAD.borders.mesc[LAD.borders.mesc$CTCF_strand == "outwards"]
LAD.borders.mesc.inwards <- LAD.borders.mesc[LAD.borders.mesc$CTCF_strand == "inwards"]
LAD.borders.mesc.ambiguous <- LAD.borders.mesc[LAD.borders.mesc$CTCF_strand == "ambiguous"]

# Prepare file names and write bed files
LAD.borders.mesc.outwards.file <- file.path(output.dir, "LAD_borders_mESC_outwards.bed")
LAD.borders.mesc.inwards.file <- file.path(output.dir, "LAD_borders_mESC_inwards.bed")
LAD.borders.mesc.ambiguous.file <- file.path(output.dir, "LAD_borders_mESC_ambiguous.bed")

export.bed(LAD.borders.mesc.outwards, LAD.borders.mesc.outwards.file)
export.bed(LAD.borders.mesc.inwards, LAD.borders.mesc.inwards.file)
export.bed(LAD.borders.mesc.ambiguous, LAD.borders.mesc.ambiguous.file)

# Prepare list of files
LAD.borders.list.stranded <- c(LAD.borders.mesc.nonctcf.file, 
                               LAD.borders.mesc.outwards.file,
                               LAD.borders.mesc.inwards.file,
                               LAD.borders.mesc.ambiguous.file)

```


### 2. Run Deeptools

```{r run deeptools}

# Run deeptools with various tracks
RunDeeptools <- function(regions.bed, tracks, expname, output.dir, labels, 
                         extend = extend, bin_size = bin_size, cores = 20) {
  
  # Prepare deeptools call
  regions.bed <- paste(regions.bed, collapse = " ")
  tracks.bw <- paste(tracks, collapse = " ")
  labels <- paste(labels, collapse = " ")
  
  output.matrix <- file.path(output.dir, 
                             paste0(expname, "-deeptoolsMatrix.gz"))
  output.table <- file.path(output.dir, 
                            paste0(expname, "-deeptoolsTable.tab"))
  
  deeptools_call <- paste("/home/t.v.schaik/mydata/miniconda3/envs/deeptools/bin/computeMatrix",
                          "reference-point",
                          "-a", format(extend, scientific = FALSE), 
                          "-b", format(extend, scientific = FALSE),
                          "-bs", format(bin_size, scientific = FALSE), 
                          "-p", cores,
                          "-R", regions.bed,
                          "-S", tracks.bw,
                          "--referencePoint", "center",
                          "--samplesLabel", labels,
                          #"--outFileNameMatrix", output.table,
                          "--outFileName", output.matrix)
  
  print(deeptools_call)
  
  # Run deeptools
  system(deeptools_call)
  
  # Return location of the table
  output.matrix
  
}

PlotHeatmap <- function(matrix, output.file, range = "1.5",
                        labels = c("border_nonctcf", "border_ctcf")) {
  
  # Prepare deeptools call
  deeptools_call <- paste("/home/t.v.schaik/mydata/miniconda3/envs/deeptools/bin/plotHeatmap",
                          "-m", matrix,
                          "-out", output.file, 
                          "-min", paste0("-", range),
                          "-max", range,
                          "--missingDataColor", "0.5",
                          "--xAxisLabel", '"CTCF distance (bp)"',
                          "--regionsLabel", paste(labels, collapse = " "), 
                          "--colorMap", "RdBu")
  
  print(deeptools_call)
  
  # Run deeptools
  system(deeptools_call)
    
}

PlotProfile <- function(matrix, output.file, 
                        labels = c("border_nonctcf", "border_ctcf")) {
  
  # Prepare deeptools call
  deeptools_call <- paste("/home/t.v.schaik/mydata/miniconda3/envs/deeptools/bin/plotProfile",
                          "-m", matrix,
                          "--perGroup",
                          "--regionsLabel", paste(labels, collapse = " "), 
                          
                          "-out", output.file)
  
  print(deeptools_call)
  
  # Run deeptools
  system(deeptools_call)
  
}

# Deeptools input
deeptools_dir <- file.path(output.dir, "deeptools")
dir.create(deeptools_dir, showWarnings = F, recursive = F)


# 1) All histone modifications
exp_name <- "LAD_borders"

track_dir <- "/DATA/scratch/usr/t.v.schaik/proj/tests/results/ts181120_pADamID_mouse/analysis_CTCF_AID/Data_NQ/ChIP_NQ/HistoneModifications/Public_2i_ChIP/"

tracks <- c("/DATA/scratch/usr/t.v.schaik/proj/3D_nucleus/results/ts180110_4DN_DataProcessing/results_mouse/tracks/normalized/bin-10kb/mESC_LMNB1-10kb-combined.bw",
            "/DATA/scratch/usr/t.v.schaik/proj/tests/results/ts181120_pADamID_mouse/analysis_CTCF_AID/Data_NQ/ChIP_NQ/CohesinFactors/2_Wapl-0D-antiCtcf_SF.3517_MQ15_sample.bw",
            "/DATA/scratch/usr/t.v.schaik/proj/tests/results/ts181120_pADamID_mouse/analysis_CTCF_AID/Data_NQ/ChIP_NQ/CohesinFactors/5_WaplC6-0h_antiRad21_sample_calibrated.bw",
            grep("H3K27me1|H3K27me2|H3K9me3", 
                 dir(track_dir, full.names = T, pattern = "E14_2i"),
                 value = T, invert = T))
            #"Data_NQ/ChIP_NQ/PRCComplex/Ring1B_2i_BR1_GSM1856442_SRR2173775_R1_MQ15_sample.bw",
            #"Data_NQ/ChIP_NQ/PRCComplex/Suz12_2i_BR1_GSM1856449_SRR2173782_R1_MQ15_sample.bw")

track_names <- c("LMNB1", "CTCF", "RAD21", 
                 str_remove(str_remove(basename(tracks[4:(length(tracks))]), "E14_2i_"), "_.*"))
                 #"Ring1B", "Suz12")

# Sort tracks and names
idx <- c(1:4, 8, 9, 10, 5, 6, 7)
tracks <- tracks[idx]
track_names <- track_names[idx]


extend <- 100000
bin_size <- 5000
deeptools_matrix <- RunDeeptools(LAD.borders.list, tracks, exp_name,
                                 deeptools_dir, track_names,
                                 extend = extend, bin_size = bin_size)

PlotHeatmap(deeptools_matrix, file.path(deeptools_dir, paste0(exp_name, "_heatmap.pdf")))
PlotProfile(deeptools_matrix, file.path(deeptools_dir, paste0(exp_name, "_profile.pdf")))


# 2) All histone modifications - stranded information 
exp_name <- "LAD_borders_CTCF_orientation"


extend <- 100000
bin_size <- 5000
deeptools_matrix <- RunDeeptools(LAD.borders.list.stranded, tracks, exp_name,
                                 deeptools_dir, track_names,
                                 extend = extend, bin_size = bin_size)

PlotHeatmap(deeptools_matrix, file.path(deeptools_dir, paste0(exp_name, "_heatmap.pdf")),
            labels = c("border_nonctcf", "border_outwards", 
                       "border_inwards", "border_ambiguous"))
PlotProfile(deeptools_matrix, file.path(deeptools_dir, paste0(exp_name, "_profile.pdf")),
            labels = c("border_nonctcf", "border_outwards", 
                       "border_inwards", "border_ambiguous"))


# 3) CTCF depletion experiments
exp_name <- "CTCF_depletion"

track_dir <- "/DATA/scratch/usr/t.v.schaik/proj/tests/results/ts181120_pADamID_mouse/analysis_CTCF_AID/ts210204_CTCF_sites_within_LADs/bigwig_scaled/"

tracks <- grep("96h", 
               dir(track_dir, full.names = T, pattern = "PT|CTCF-EL"),
               value = T, invert = T)

# Sort tracks and names
idx <- c(4, 1, 3, 2)
tracks <- tracks[idx]
track_names <- c("PT_0h", "CTCFEL_0h", "CTCFEL_6h", "CTCFEL_24h")


extend <- 100000
bin_size <- 5000
deeptools_matrix <- RunDeeptools(LAD.borders.list, tracks, exp_name,
                                 deeptools_dir, track_names,
                                 extend = extend, bin_size = bin_size)

PlotHeatmap(deeptools_matrix, file.path(deeptools_dir, paste0(exp_name, "_heatmap.pdf")))
PlotProfile(deeptools_matrix, file.path(deeptools_dir, paste0(exp_name, "_profile.pdf")))

```

### 3. Plot in R

Let's make a better plot in R.

```{r plot in R, fig.width = 7, fig.height = 8}

# Load deeptools output
bins <- 1:(extend*2/bin_size)

tib <- read_tsv(file.path(output.dir, "deeptools/LAD_borders-deeptoolsMatrix.gz"), 
                skip = 1, 
                col_names = c("seqnames", "start", "end", 
                              "name", "unknown", "strand",
                              paste(rep(track_names, 
                                        each = length(bins)), 
                                    bins, sep = "_")))

tib_gather <- tib %>%
  dplyr::select(-seqnames, -start, -end, -name, -unknown, -strand) %>%
  add_column(border = 1:nrow(.),
             ctcf = rep(c("nonctcf", "ctcf"), 
                        times = c(length(LAD.borders.mesc.nonctcf),
                                  length(LAD.borders.mesc.ctcf)))) %>%
  gather(key, value, -border, -ctcf) %>%
  separate(key, c("track", "bin"), remove = T) %>%
  mutate(track = factor(track, levels = track_names),
         bin = as.numeric(bin),
         bin = (bin * bin_size) - (extend + 0.5 * bin_size),
         bin = bin / 1e3) %>%
  drop_na()

# Fix outliers by setting them to a limit
limit <- quantile(tib_gather$value, 0.999)
tib_gather$value[tib_gather$value > limit] <- limit

# Filter borders overlapping with active genes
idx_active <- as_tibble(mcols(c(LAD.borders.mesc.nonctcf,
                                  LAD.borders.mesc.ctcf))) %>%
  pull(ovl_gene)

# Plot
tib_gather %>%
  filter(border %in% which(! idx_active)) %>%
  drop_na() %>%
  ggplot(aes(x = bin, y = value, col = ctcf, fill = ctcf,
             group = interaction(ctcf, track))) +
  annotate("rect", xmin = 0, xmax = 1e3, ymin = -Inf, ymax = Inf, 
           fill = "grey", alpha = 0.3) +  
  geom_hline(yintercept = 0, col = "black", linetype = "dashed") +
  geom_vline(xintercept = 0, col = "black", linetype = "dashed") +
  stat_summary(fun = mean, geom = "line", size = 1) +
  stat_summary(fun.data = mean_se, geom = "ribbon", alpha = 0.25, col = NA,
               fun.args = list(mult = 1.96)) +
  facet_wrap(~ track, scales = "free_y", ncol = 2) +
  xlab("Distance from LAD border (kb)") +
  ylab("Score") +
  coord_cartesian(xlim = c(-extend/1e3, extend/1e3)) +
  scale_fill_brewer(palette = "Set2", name = "Border class") +
  scale_color_brewer(palette = "Set2", name = "Border class") +
  theme_bw() +
  theme(aspect.ratio = 1, 
        axis.text.x = element_text(angle = 90, hjust = 1))

# tib_gather %>%
#   mutate(value = case_when(value > 2 ~ 2,
#                            value < -2 ~ -2,
#                            T ~ value)) %>%
#   ggplot(aes(x = bin, y = border, fill = value)) +
#   geom_tile() +
#   facet_grid(. ~ track) +
#   xlab("Distance from LAD border (kb)") +
#   ylab("Score") +
#   scale_fill_gradient2(low = "red", mid = "white", high = "blue",
#                        na.value = "grey50") +
#   theme_bw() +
#   theme(axis.text.x = element_text(angle = 90, hjust = 1))




# Repeat for CTCF orientation
# Load deeptools output
bins <- 1:(extend*2/bin_size)

tib <- read_tsv(file.path(output.dir, "deeptools/LAD_borders_CTCF_orientation-deeptoolsMatrix.gz"), 
                skip = 1, 
                col_names = c("seqnames", "start", "end", 
                              "name", "unknown", "strand",
                              paste(rep(track_names, 
                                        each = length(bins)), 
                                    bins, sep = "_")))

tib_gather <- tib %>%
  dplyr::select(-seqnames, -start, -end, -name, -unknown, -strand) %>%
  add_column(border = 1:nrow(.),
             ctcf = rep(c("nonctcf", "outwards", "inwards", "ambiguous"), 
                        times = c(length(LAD.borders.mesc.nonctcf),
                                  length(LAD.borders.mesc.outwards),
                                  length(LAD.borders.mesc.inwards),
                                  length(LAD.borders.mesc.ambiguous)))) %>%
  gather(key, value, -border, -ctcf) %>%
  separate(key, c("track", "bin"), remove = T) %>%
  mutate(track = factor(track, levels = track_names),
         bin = as.numeric(bin),
         bin = (bin * bin_size) - (extend + 0.5 * bin_size),
         bin = bin / 1e3,
         ctcf = factor(ctcf, c("outwards", "inwards", "ambiguous", "nonctcf"))) %>%
  drop_na()

# Fix outliers by setting them to a limit
limit <- quantile(tib_gather$value, 0.999)
tib_gather$value[tib_gather$value > limit] <- limit

# Filter borders overlapping with active genes
idx_active <- as_tibble(mcols(c(LAD.borders.mesc.nonctcf,
                                LAD.borders.mesc.outwards,
                                LAD.borders.mesc.inwards,
                                LAD.borders.mesc.ambiguous))) %>%
  pull(ovl_gene)

# Plot
tib_gather %>%
  filter(border %in% which(! idx_active)) %>%
  drop_na() %>%
  ggplot(aes(x = bin, y = value, col = ctcf, fill = ctcf,
             group = interaction(ctcf, track))) +
  annotate("rect", xmin = 0, xmax = 1e3, ymin = -Inf, ymax = Inf, 
           fill = "grey", alpha = 0.2) +  
  geom_hline(yintercept = 0, col = "black", linetype = "dashed") +
  geom_vline(xintercept = 0, col = "black", linetype = "dashed") +
  stat_summary(fun = mean, geom = "line", size = 1) +
  stat_summary(fun.data = mean_se, geom = "ribbon", alpha = 0.1, col = NA,
               fun.args = list(mult = 1.96)) +
  facet_wrap(~ track, scales = "free_y", ncol = 2) +
  xlab("Distance from LAD border (kb)") +
  ylab("Score") +
  coord_cartesian(xlim = c(-extend/1e3, extend/1e3)) +
  scale_fill_brewer(palette = "Set2", name = "Border class") +
  scale_color_brewer(palette = "Set2", name = "Border class") +
  theme_bw() +
  theme(aspect.ratio = 1, 
        axis.text.x = element_text(angle = 90, hjust = 1))

```



### Conclusion




### SessionInfo

```{r sessioninfo}

sessionInfo()

```











