---
title: "DamID changes versus LAD size and score"
author: "Tom van Schaik"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document:
    theme: journal #cerulean
    highlight: monochrome
    toc: true
    toc_float: true
    code_folding: show
  editor_options:
    chunk_output_type: console
---


### Project description

Effect of cohesin looping on genome architecture, from the perspective of the 
nuclear lamina.


### Introduction

In a previous document, I classified LAD borders as 1) CTCF-bound or not 2) 
shared or unique. Here, I will visualize what the effect is of CTCF depletion on
the entire LADs.

To also capture newly-formed LADs after a certain treatment, I will first 
determine a union set of all LAD models. This consensus model will be used 
later to correlate with external features.


### Method

Prepare consensus LAD model based on individual HMM models.

Load (z-scale) DamID tracks and plot effect on different types of LAD borders.
To do this, determine the average score of a LAD. This is a very robust 
estimation compared to individual 10 kb bins.


### Set-up

Load the libraries and set the parameters.

```{r set-up}

# Load dependencies
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(GenomicRanges))
suppressPackageStartupMessages(library(rtracklayer))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(UpSetR))
suppressPackageStartupMessages(library(ggbeeswarm))
suppressPackageStartupMessages(library(caTools))
suppressPackageStartupMessages(library(GGally))
suppressPackageStartupMessages(library(ggrastr))

# Prepare output 
output.dir <- "ts220113_DamID_changes_versus_LAD_size_and_score"
dir.create(output.dir, showWarnings = FALSE)

```

Prepare knitr output.

```{r knits setup}
library(knitr)
opts_chunk$set(fig.width = 10, fig.height = 4, 
               message = F, warning = F,
               dev=c('png', 'pdf'), fig.path = file.path(output.dir, "figures/")) 
pdf.options(useDingbats = FALSE)
```

List functions.

```{r functions}

LoadHMM <- function(metadata, hmm.dir, black.list = NULL, min.size = 5e4) {
  # Load LAD calls
  
  # Initiate GRangesList
  grlist.hmm <- GRangesList()
  # Get the samples names
  sample.names <- metadata$HMM
  
  # Load the data
  for (s in sample.names) {
    gr <- import(file.path(hmm.dir, s))
    # setdiff with black list (i.e. centromeres and NA-stretches)
    if (! is.null(black.list)) gr <- GenomicRanges::setdiff(gr, black.list)
    # Filter for size
    gr <- gr[width(gr) > min.size]
    grlist.hmm <- c(grlist.hmm, GRangesList(gr))
  }
  names(grlist.hmm) <- metadata$sample
  
  grlist.hmm
  
}

OverlapWithBins <- function(gr, gr.list) {
  # Overlap LADs with bins
  
  mcols(gr) <- NULL
  
  for (s in names(gr.list)) {
    tmp <- gr.list[[s]]
    mcols(gr)[, s] <- factor(ifelse(overlapsAny(gr, tmp),
                                    "LAD", "iLAD"),
                             levels = c("iLAD", "LAD"))
  }
  gr
  
}

LongNAStretches <- function(gr) {
  # Find long stretches of NAs, I don't want these as domains
  t_nas <- as_tibble(mcols(gr)) %>% 
    mutate_all(~!is.na(.)) %>%
    add_column(bin = 1:length(gr)) %>%
    gather(key, value, -bin) %>%
    group_by(bin) %>%
    summarise(non_na = sum(value))
  
  rle_nas <- rle(t_nas$non_na == 0)
  t_nas <- t_nas %>%
    add_column(length = rep(rle_nas$lengths, rle_nas$lengths),
               value = rep(rle_nas$values, rle_nas$lengths)) %>%
    filter(length > 10, value == T)
  
  na_stretches <- gr
  mcols(na_stretches) <- NULL
  na_stretches <- GenomicRanges::reduce(na_stretches[t_nas$bin])
  
  na_stretches
}

ConsensusLADModel <- function(LADs.gr, samples) {
  # Get a consensus LAD model of the samples
  
  # Get bins that are classified as LAD in at least one sample
  idx <- as_tibble(mcols(LADs.gr)) %>%
    dplyr::select(samples) %>%
    mutate(idx = rowSums(. == "LAD") > 0) %>%
    pull(idx)
  gr <- LADs.gr[idx]
  
  # Reduce to one LAD model
  gr <- reduce(gr)
  
  gr
  
}

LADScores <- function(damid, gr, samples) {
  # Calculate mean scores per LAD, given a LAD model
  
  # Find overlaps
  ovl <- findOverlaps(damid, gr, select = "first")
  
  # Calculation
  tib <- as_tibble(mcols(damid)) %>%
    dplyr::select(samples) %>%
    add_column(LAD = ovl) %>%
    drop_na(LAD) %>%
    group_by(LAD) %>%
    summarise_all(mean, na.rm = T) %>%
    dplyr::select(-LAD)
  
  mcols(gr) <- tib
  gr
  
}

PlotLADScores <- function(gr, title, to.plot = "size") {
  # Plot changes in LAD score versus "to.plot" parameter (size or mean score)
  
  # Prepare tib
  tib <- as_tibble(gr) %>%
    rename_at(vars(ends_with("h")), str_replace, ".*_", "t_") %>%
    mutate(mean_score = rowMeans(dplyr::select(., starts_with("t_")))) %>%
    rowwise() %>%
    mutate(diff_6h = ifelse("t_6h" %in% names(.),
                            t_6h - t_0h, NA),
           diff_24h = ifelse("t_24h" %in% names(.),
                            t_24h - t_0h, NA),
           diff_96h = ifelse("t_96h" %in% names(.),
                            t_96h - t_0h, NA)) %>%
    ungroup() %>%
    dplyr::select(-starts_with("t_")) %>%
    gather(key, value, starts_with("diff_")) %>%
    mutate(key = factor(key, levels = c("diff_6h", "diff_24h", "diff_96h")))
  
  # Plot
  plt <- tib %>%
    ggplot() +
      facet_grid(. ~ key) +
      ggtitle(title) +
      ylab("DamID change with t=0h (z-score)") +
      theme_bw() + 
      theme(aspect.ratio = 1)
  
  if (to.plot == "size") {
    plt <- plt +
      geom_point(aes(x = width / 1e6, y = value), alpha = 0.3) +
      xlab("LAD size (Mb)") +
      scale_x_log10()
  } else if (to.plot == "score") {
    plt <- plt +
      geom_point(aes(x = mean_score, y = value), alpha = 0.3) +
      xlab("Mean LAD score (z-score)")
  }
  
  plt +
    geom_hline(yintercept = 0, col = "red", linetype = "dashed")
  
}

# From Fede:
# ggpairs custom functions
corColor <- function(data, mapping, color = I("black"), sizeRange = c(1, 3), ...) {

  x   <- eval_data_col(data, mapping$x)
  y   <- eval_data_col(data, mapping$y)
  r   <- cor(x, y)
  rt  <- format(r, digits = 3)
  tt  <- as.character(rt)
  cex <- max(sizeRange)

  # helper function to calculate a useable size
  percent_of_range <- function(percent, range) {
    percent * diff(range) + min(range, na.rm = TRUE)
  }

  # plot correlation coefficient
  p <- ggally_text(label = tt, mapping = aes(), xP = 0.5, yP = 0.5,
                   # size = I(percent_of_range(cex * abs(r), sizeRange)), 
                   size = 6, 
                   color = color, ...) +
    theme(panel.grid.minor=element_blank(),
          panel.grid.major=element_blank())

  corColors <- RColorBrewer::brewer.pal(n = 7, name = "RdYlBu")[2:6]

  if (r <= boundaries[1]) {
    corCol <- corColors[1]
  } else if (r <= boundaries[2]) {
    corCol <- corColors[2]
  } else if (r < boundaries[3]) {
    corCol <- corColors[3]
  } else if (r < boundaries[4]) {
    corCol <- corColors[4]
  } else {
    corCol <- corColors[5]
  }

  p <- p +
    theme(panel.background = element_rect(fill = corCol))

  return(p)
}

customScatter <- function (data, mapping) 
{
    p <- ggplot(data = data, mapping) + 
      geom_point() +
      geom_smooth(method = "lm", se = T, col = "red") +
      theme_bw()
    
    p 
}

```


### 1. Load data

Load the required data. 

#### 1.1 Previous objects

First, simply objects from previous documents.

```{r load previous data, cache = T}

# Read .rds files
input.dir <- "ts220113_CTCF_enrichment_at_LAD_borders"
metadata.cells <- readRDS(file.path(input.dir, "metadata.rds"))
LADs.cells <- readRDS(file.path(input.dir, "LADs_pA.rds"))
CTCF.sites <- readRDS(file.path(input.dir, "CTCF_sites_pA.rds"))

input.dir <- "ts220113_effect_of_CTCF_depletion_on_LAD_borders"
metadata.damid <- readRDS(file.path(input.dir, "metadata_damid.rds"))
bin.size <- readRDS(file.path(input.dir, "bin_size.rds"))
damid <- readRDS(file.path(input.dir, "damid.rds"))

```

#### 1.2 LADs

I also need LADs for this. I will load the combined calls. Normally, I make sure
that LADs do not overlap complete centromeres. Of course, this is not an issue
with mouse cells where centromeres are at the beginning of the chromosomes. I 
will remove long NA-stretches of unmappable DNA. 

First, determine these NA stretches.

```{r NA-stretches, cache = T}

# Define NA-stretches
na.stretches <- LongNAStretches(damid)

```

Then, load LADs and remove NA-stretches.

```{r load lads, cache = T, warning = F, message = F}

# The directory with files
hmm.dir <- file.path("../results_NQ/HMM/", paste0("bin-", bin.size))

# Add the HMM files to metadata.damid
metadata.damid <- metadata.damid %>%
  mutate(HMM = str_replace(file, ".norm.txt.gz", "_AD.bed.gz")) #%>%
  #filter(timepoint != "96h")

# Load LADs - using .bed files and not overlapping NA-stretches
LADs.grlist <- LoadHMM(metadata.damid, hmm.dir, black.list = na.stretches)

# Get per-bin information
LADs.gr <- OverlapWithBins(damid, LADs.grlist)

```


### 2. Overlap between LADs

What is the overlap between LADs in the different samples? Comparisons to make:

  * Between 0h timepoints.
  * Between same condition.

```{r overlap in LADs, cache = T, fig.width = 6, fig.height = 4.5}

# Plot overlap between LADs
PlotLADOverlap <- function(gr, samples) {
  #
  tib <- as_tibble(mcols(gr)) %>%
    dplyr::select(samples)
  
  # Convert to binary data.frame
  df <- data.frame(tib == "LAD") + 0
  
  # Plot 
  upset(df, order.by = "freq", nintersects = 10, sets = samples, keep.order = T)
}

PlotLADOverlap(LADs.gr, samples = c("PT_0h", "CTCFEL_0h", "WAPL_0h", 
                                    "CTCFWAPL_0h", "RAD21_0h"))

PlotLADOverlap(LADs.gr, samples = c("PT_0h", "PT_24h"))
PlotLADOverlap(LADs.gr, samples = c("CTCFEL_0h", "CTCFEL_6h", 
                                    "CTCFEL_24h"))
PlotLADOverlap(LADs.gr, samples = c("WAPL_0h", "WAPL_6h", 
                                    "WAPL_24h"))
PlotLADOverlap(LADs.gr, samples = c("CTCFWAPL_0h", "CTCFWAPL_6h", 
                                    "CTCFWAPL_24h"))
PlotLADOverlap(LADs.gr, samples = c("RAD21_0h", "RAD21_6h", 
                                    "RAD21_24h"))

```

Overall, this should show that LADs do not change that much during the depletion
of cohesin looping factors. Small details can be seen. For instance.

  * Slightly fewer LADs in the WAPL-AID cell line, probably due to stronger CTCF
    escaping.
  * Decreasing LADs upon CTCF depletion, probably due to small LADs 
    disappearing.
  * Decreasing LADs upon WAPL depletion, probably due to stronger CTCF escaping.
  * Increasing(!) LADs upon double depletion, probably due to increasing 
    interactions of weak LADs.
    

### 3. LAD size and strength versus LAD change

Here, I want to reproduce a previous figure that shows that strong LADs become
stronger upon CTCF depletion while weak LADs become weaker. This will be 
extended with the other AID cell lines.

I should decide what definition of LADs to use though. For now, let's do it as 
before: a union set of all time points (to also capture increasing LADs).

First, I will calculate LAD scores.

```{r calculate LAD scores, cache = T, fig.width = 8, fig.height = 3.5}

# Get LAD definition per condition
LADs.list.condition <- lapply(levels(metadata.damid$condition),
                              function(x) {
                                samples <- metadata.damid %>%
                                  filter(condition == x &
                                           timepoint != "96h") %>%
                                  pull(sample)
                                samples <- as.character(samples)
                                ConsensusLADModel(LADs.gr, samples)
                              })
names(LADs.list.condition) <- levels(metadata.damid$condition)



# Instead, prepare one consensus model that every condition uses,
# save individual models
LADs.list.condition.individual <- LADs.list.condition

LADs.consensus.gr <- LADs.gr

mcols(LADs.consensus.gr) <- tibble(PT = overlapsAny(LADs.consensus.gr, 
                                                    LADs.list.condition[["PT"]]),
                                   CTCFEL = overlapsAny(LADs.consensus.gr, 
                                                        LADs.list.condition[["CTCFEL"]]),
                                   RAD21 = overlapsAny(LADs.consensus.gr, 
                                                       LADs.list.condition[["RAD21"]]),
                                   WAPL = overlapsAny(LADs.consensus.gr, 
                                                      LADs.list.condition[["WAPL"]]),
                                   CTCFWAPL = overlapsAny(LADs.consensus.gr, 
                                                          LADs.list.condition[["CTCFWAPL"]])) %>%
  mutate_all(function(x) ifelse(x, "LAD", "iLAD"))


PlotLADOverlap(LADs.consensus.gr, samples = c("PT", "CTCFEL", "RAD21", 
                                              "WAPL", "CTCFWAPL"))


# Get LAD definition per condition
LADs.consensus <- ConsensusLADModel(LADs.consensus.gr, 
                                    c("PT", "CTCFEL", "RAD21", "WAPL", "CTCFWAPL"))



# Update individual LADs with the consensus model
LADs.list.condition <- rep(list(LADs.consensus),
                           length(levels(metadata.damid$condition)))
names(LADs.list.condition) <- levels(metadata.damid$condition)






# Get LAD score for every condition
LADs.list.condition.individual <- lapply(levels(metadata.damid$condition),
                                         function(x) {
                                           samples <- metadata.damid %>% 
                                             filter(condition == x) %>%
                                             pull(sample)
                                           samples <- as.character(samples)
                                           gr <- LADs.list.condition.individual[[x]]
                                           LADScores(damid, gr, samples)})
names(LADs.list.condition.individual) <- levels(metadata.damid$condition)

LADs.list.condition <- lapply(levels(metadata.damid$condition),
                              function(x) {
                                samples <- metadata.damid %>% 
                                  filter(condition == x) %>%
                                  pull(sample)
                                samples <- as.character(samples)
                                gr <- LADs.list.condition[[x]]
                                LADScores(damid, gr, samples)})
names(LADs.list.condition) <- levels(metadata.damid$condition)

lapply(levels(metadata.damid$condition),
       function(x) PlotLADScores(LADs.list.condition[[x]], x, "size"))

lapply(levels(metadata.damid$condition),
       function(x) PlotLADScores(LADs.list.condition[[x]], x, "score"))


# Get LAD score for every condition
samples <- metadata.damid %>% 
  filter(condition %in% c("PT", "CTCFEL", "RAD21", "WAPL", "CTCFWAPL")) %>%
  pull(sample)
samples <- as.character(samples)

LADs.consensus <- LADScores(damid, LADs.consensus, 
                            samples)

 
```

Looks good. Note that these are still exploratory figures, and later analyses
will formalize these results.

I also want a plot that shows the absolute differences for every depletion
experiment.

```{r differences per condition, cache = T, fig.width = 8, fig.height = 3.5, message = F, warning = F}

# Get absolute differences per condition
DifferenceWithZero <- function(gr) {
  tib <- as_tibble(mcols(gr)) %>%
    rename_at(1, ~ c("t_0h")) %>%
    mutate_at(2:ncol(.), function(x) x - .$t_0h) %>%
    dplyr::select(-1) %>%
    gather(key, value)
  tib
}

tib <- purrr::reduce(lapply(LADs.list.condition, 
                            DifferenceWithZero),
                     bind_rows) %>%
  separate(key, c("condition", "timepoint")) %>%
  mutate(condition = factor(condition, levels = levels(metadata.damid$condition)),
         timepoint = factor(timepoint, levels = levels(metadata.damid$timepoint)))

# And plot
tib %>%
  ggplot(aes(x = condition, y = value, col = condition)) +
  geom_quasirandom() +
  geom_boxplot(fill = NA, outlier.shape = NA, col = "black") +
  geom_hline(yintercept = 0, col = "blue", linetype = "dashed") +
  facet_grid(. ~ timepoint) +
  xlab("") +
  ylab("LAD difference (z-score)") +
  theme_bw() +
  theme(aspect.ratio = 1,
        axis.text.x = element_text(angle = 90, hjust = 1))

tib %>%
  filter(! condition %in% c("CTCFNQ", "CTCF"),
         timepoint != "96h") %>%
  ggplot(aes(x = condition, y = value, col = condition)) +
  geom_quasirandom() +
  geom_boxplot(fill = NA, outlier.shape = NA, col = "black") +
  geom_hline(yintercept = 0, col = "blue", linetype = "dashed") +
  facet_grid(. ~ timepoint) +
  xlab("") +
  ylab("LAD difference (z-score)") +
  scale_color_brewer(palette = "Dark2") +
  theme_bw() +
  theme(aspect.ratio = 1,
        axis.text.x = element_text(angle = 90, hjust = 1))

tib %>%
  filter(! condition %in% c("CTCFNQ", "CTCF"),
         timepoint != "96h") %>%
  ggplot(aes(x = condition, y = abs(value), col = condition)) +
  geom_quasirandom() +
  geom_boxplot(fill = NA, outlier.shape = NA, col = "black") +
  geom_hline(yintercept = 0, col = "blue", linetype = "dashed") +
  facet_grid(. ~ timepoint) +
  xlab("") +
  ylab("LAD difference (absolute z-score)") +
  scale_color_brewer(palette = "Dark2") +
  theme_bw() +
  theme(aspect.ratio = 1,
        axis.text.x = element_text(angle = 90, hjust = 1))


# Include st-dev in the plot
sd_fun <- function(x){
    return(data.frame(y = 1.25, label = round(sd(x), 2)))
}

tib %>%
    filter(! condition %in% c("CTCFNQ", "CTCF"),
           timepoint != "96h") %>%
    ggplot(aes(x = condition, y = value, col = condition)) +
    geom_quasirandom() +
    geom_boxplot(fill = NA, outlier.shape = NA, col = "black") +
    stat_summary(fun.data = sd_fun, geom = "text", col = "black") +
    geom_hline(yintercept = 0, col = "blue", linetype = "dashed") +
    facet_grid(. ~ timepoint) +
    xlab("") +
    ylab("LAD difference (z-score)") +
    scale_color_brewer(palette = "Dark2") +
    theme_bw() +
    theme(aspect.ratio = 1,
          axis.text.x = element_text(angle = 90, hjust = 1))
 
```

Very clear. Nicely summarizes that CTCF/WAPL depletion has the biggest 
genome-wide effects.


### 4. Plot some example tracks 

To explain these steps, prepare some example tracks that can be included in a 
manuscript.

```{r example tracks, cache = T, fig.width = 8, fig.height = 6}

PlotDataTracks <- function(gr_norm_mean, name, samples,
                           gr_regions,
                           centromeres = NULL,
                           plot_chr = "chr1", 
                           plot_start = 1, plot_end = 100e6,
                           smooth = 1, fill = "cell",
                           filter_samples = NULL,
                           free_y = T) {
  
  ################################
  ## Prepare bin scores
  
  # Get the scores for the samples
  tib <- as_tibble(gr_norm_mean) %>%
    dplyr::select(seqnames, start, end, 
                  contains("PT"),
                  matches(name))
  
  if (! is.null(filter_samples)) {
    tib <- tib %>%
      dplyr::select(-matches(paste(filter_samples, collapse = "|")))
  }
  
  if (ncol(tib) != 5) {
    stop("Works with 2 columns only")
  }
  
  if (smooth > 1) {
    tib <- tib %>%
      mutate_at(vars(contains("_")), 
                runmean, k = smooth)
  }
  
  # Filter for plotting window
  tib <- tib %>%
    filter(seqnames == plot_chr,
           start >= plot_start,
           end <= plot_end)
  
  tib_names <- names(tib)[4:5]
  
  # Gather
  tib_gather <- tib %>% 
    rename_at(4:5, ~c("control", "treatment")) %>%
    gather(key, value, -seqnames, -start, -end) %>%
    mutate(class = "bin")
  
  ################################
  ## Prepare region scores
  tib_regions <- as_tibble(gr_regions) %>%
    dplyr::select(seqnames, start, end, 
                  contains("PT"),
                  matches(name))
  
  if (! is.null(filter_samples)) {
    tib_regions <- tib_regions %>%
      dplyr::select(-matches(paste(filter_samples, collapse = "|")))
  }
  
  # Filter for plotting window
  tib_regions <- tib_regions %>%
    filter(seqnames == plot_chr,
           start < plot_end,
           end > plot_start) %>%
    rename_at(4:5, ~c("control", "treatment")) %>%
    mutate(diff = treatment - control)
  
  
  tib_regions_gather <- tib_regions %>%
    rename_at(4:5, ~c("control", "treatment")) %>%
    gather(key, value, -seqnames, -start, -end) %>%
    mutate(class = "region")
  
  
  ################################
  ## Combine into one tibble
  tib_combined <- bind_rows(tib_gather,
                            tib_regions_gather) %>%
    drop_na() %>%
    mutate(key = factor(key, c("control", "treatment", "diff")),
           interaction = interaction(key, class),
           interaction = factor(interaction, unique(interaction)))
  
  
  # Limits
  #ylimits_bin <- quantile(tib_gather$value, c(0, 1), na.rm = T)
  ylimits_bin <- c(-2, 2)
  tib_ylimits <- bind_rows(tibble(xmin = tib_gather$start[1]/1e6, 
                                  xmax = tib_gather$start[1]/1e6 + 1,
                                  ymin = ylimits_bin[1], ymax = ylimits_bin[2],
                                  interaction = c("control.bin", "treatment.bin",
                                                  "control.region", "treatment.region")),
                           tibble(xmin = tib_gather$start[1]/1e6, 
                                  xmax = tib_gather$start[1]/1e6 + 1,
                                  ymin = -1, ymax = 1,
                                  interaction = c("diff.region"))) %>%
    mutate(interaction = factor(interaction, levels(tib_combined$interaction)))
  
  # Remove values higher than the limits
  tib_combined <- tib_combined %>%
    mutate(value = case_when(value < ylimits_bin[1] ~ ylimits_bin[1],
                             value > ylimits_bin[2] ~ ylimits_bin[2],
                             T ~ value))
  
  # Colors
  plt <- tib_combined %>%
    ggplot(aes(fill = key)) +
    rasterize(geom_rect(aes(xmin = start / 1e6, xmax = end / 1e6, 
                            ymin = 0, ymax = value)),
              dpi = 300) +
    geom_rect(data = tib_ylimits, 
              aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
              fill = NA, col = NA) +
    geom_hline(yintercept = 0, size = 0.5) +
    coord_cartesian(xlim = c(max(c(plot_start,
                                   min(tib_combined$start))) / 1e6,
                             min(c(plot_end,
                                   max(tib_combined$end))) / 1e6)) +
    xlab(paste0(plot_chr, " (Mb)")) +
    ylab("pA-DamID (z-score)") +
    ggtitle(paste(tib_names[2], "vs", tib_names[1])) +
    scale_x_continuous(expand = c(0, 0)) + 
    scale_y_continuous(expand = c(0, 0)) +
    scale_fill_brewer(palette = "Set1", guide = F) +
    theme_classic() +
    theme(aspect.ratio = 0.15)
  
  if (free_y) {
    plt <- plt + 
      facet_grid(interaction ~ ., scales = "free_y")
  } else {
    plt <- plt + 
      facet_grid(interaction ~ .)
  }
  
  plot(plt)
  
}

# Plot some example tracks
PlotDataTracks(gr_norm_mean = damid,
               name = "PT",
               samples = metadata.damid,
               gr_regions = LADs.list.condition$PT,
               plot_chr = "chr3", plot_start = 0, plot_end = 1e9, 
               smooth = 9)

PlotDataTracks(gr_norm_mean = damid,
               name = "CTCFEL",
               samples = metadata.damid,
               gr_regions = LADs.list.condition$CTCFEL,
               plot_chr = "chr3", plot_start = 0, plot_end = 1e9, 
               smooth = 9,
               filter_samples = c("PT", "_6h", "_96h"))

PlotDataTracks(gr_norm_mean = damid,
               name = "RAD21",
               samples = metadata.damid,
               gr_regions = LADs.list.condition$RAD21,
               plot_chr = "chr3", plot_start = 0, plot_end = 1e9, 
               smooth = 9,
               filter_samples = c("PT", "_6h", "_96h"))

PlotDataTracks(gr_norm_mean = damid,
               name = "^WAPL",
               samples = metadata.damid,
               gr_regions = LADs.list.condition$WAPL,
               plot_chr = "chr3", plot_start = 0, plot_end = 1e9, 
               smooth = 9,
               filter_samples = c("PT", "_6h", "_96h"))

PlotDataTracks(gr_norm_mean = damid,
               name = "CTCFWAPL",
               samples = metadata.damid,
               gr_regions = LADs.list.condition$CTCFWAPL,
               plot_chr = "chr3", plot_start = 0, plot_end = 1e9, 
               smooth = 9,
               filter_samples = c("PT", "_6h", "_96h"))


# gr_norm_mean = damid; 
# name = "CTCFEL"; 
# samples = metadata.damid;
# gr_regions = LADs.list.condition$CTCFEL
# centromeres = NULL;
# plot_chr = "chr1";
# plot_start = 1; plot_end = 100e6;
# smooth = 1; fill = "cell";
# filter_samples = NULL;
# free_y = T
 
```

Simple, but convincing. Good.


### Save data

```{r save data}

export.bed(LADs.list.condition.individual$PT, file.path(output.dir, "LADs_PT.bed"))
export.bed(LADs.list.condition.individual$CTCF, file.path(output.dir, "LADs_CTCF.bed"))
export.bed(LADs.list.condition.individual$WAPL, file.path(output.dir, "LADs_WAPL.bed"))
export.bed(LADs.list.condition.individual$CTCFWAPL, file.path(output.dir, "LADs_CTCFWAPL.bed"))
export.bed(LADs.list.condition.individual$RAD21, file.path(output.dir, "LADs_RAD21.bed"))

export.bed(LADs.consensus, file.path(output.dir, "LADs_consensus.bed"))

saveRDS(LADs.list.condition, file.path(output.dir, "LADs_list.rds"))
saveRDS(LADs.list.condition.individual, file.path(output.dir, "LADs_list_individual.rds"))
saveRDS(LADs.consensus, file.path(output.dir, "LADs_consensus.rds"))

# Also, save bigwig files of differences
lad.difference.dir <- file.path(output.dir, "bigwig_lad_difference")
dir.create(lad.difference.dir, showWarnings = FALSE)

chr.sizes <- read.table("~/mydata/data/genomes/mm10/mm10.chrom.sizes", sep = "\t")
row.names(chr.sizes) <- chr.sizes[, 1]

timepoints <- c("0h", "6h", "24h", "96h")
for (l in names(LADs.list.condition)) {
  gr <- LADs.list.condition[[l]]
  for (i in 2:ncol(mcols(gr))) {
    tmp <- gr
    mcols(tmp) <- data.frame(score = mcols(gr)[, i] - mcols(gr)[, 1])
    tmp <- tmp[! is.na(tmp$score)]
    seqlengths(tmp) <- chr.sizes[seqlevels(tmp), 2]
    export.bw(tmp, file.path(lad.difference.dir, paste0(l, "_",
                                                        timepoints[i], 
                                                        "_diff.bw")))
  }
}

```


### Conclusion

I prepared a consensus LAD model and quantified the mean LaminB1 scores on 
these LADs. In another document, I will use these data to correlate with 
LAD features to determine which LADs are affected by the protein depletions.


### SessionInfo

```{r sessioninfo}

sessionInfo()

```











