---
title: "Replicate statistics"
author: "Tom"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document:
    theme: journal #cerulean
    highlight: monochrome
    toc: true
    toc_float: true
    code_folding: show
  editor_options:
    chunk_output_type: console
---

```{r setup}

```

### Overview

This document provides quality control figures for DamID replicates.

### Set-up 

```{r temporary test set-up, include = FALSE} 

# # # Libraries used
# library(GenomicRanges)
# library(GenomicAlignments)
# library(rtracklayer)
# library(ggplot2)
# library(reshape2)
# library(GGally)
# library(RColorBrewer)
# 
# # Basename
# basename <- "H1_LMNB1"
# samples <- c("H1_r1_LMNB1", "H1_r2_LMNB1", "H1_r3_LMNB1")
# 
# # Bin sizes
# bins <- c("5", "20", "80")
# bins.kb <- paste0(bins, "kb")
# bins.with_gatc <- c("gatc", paste0(bins, "kb"))
# 
# bins.count <- length(bins.with_gatc)
# 
# # Input directories
# map_dir <- "/DATA/usr/t.v.schaik/proj/3D_nucleus/results/ts180110_4DN_DataProcessing/results/mapped"
# counts_dir <- "/DATA/usr/t.v.schaik/proj/3D_nucleus/results/ts180110_4DN_DataProcessing/results/counts"
# norm_dir <- "/DATA/usr/t.v.schaik/proj/3D_nucleus/results/ts180110_4DN_DataProcessing/results/normalized"
# hmm_dir <- "/DATA/usr/t.v.schaik/proj/3D_nucleus/results/ts180110_4DN_DataProcessing/results/HMM"

```

```{r damid set-up}

# Libraries used
library(GenomicRanges)
library(GenomicAlignments)
library(rtracklayer)
library(ggplot2)
library(reshape2)
library(GGally)
library(RColorBrewer)

# Basenames
basename <- snakemake@params[["basename"]]
samples <- snakemake@config[["replicates"]][[basename]]

if (length(samples) < 2) {
  stop("Not enough samples!")
} 

# Bin sizes
bins <- snakemake@config[["bins"]]
bins <- strsplit(bins, ",")[[1]]
bins.kb <- paste0(bins, "kb")
bins.with_gatc <- c("gatc", paste0(bins, "kb"))

bins.count <- length(bins.with_gatc)

# Input directories - note that the .Rmd file is the working directory!
map_dir <- file.path(snakemake@config[["report_relative_location"]],
                     snakemake@config[["output_dir"]],
                     snakemake@config["out_map"])
counts_dir <- file.path(snakemake@config[["report_relative_location"]],
                        snakemake@config[["output_dir"]],
                        snakemake@config[["out_count"]])
norm_dir <- file.path(snakemake@config[["report_relative_location"]],
                      snakemake@config[["output_dir"]],
                      snakemake@config[["out_norm"]])
hmm_dir <- file.path(snakemake@config[["report_relative_location"]],
                     snakemake@config[["output_dir"]],
                     snakemake@config[["out_hmm"]])

```

Samples to be analyzed:

```{r damid set-up printed}
print(basename)

print(samples)

```

### Read-in data

I will read in the data

  * Normalized scores
  * HMM model

```{r read-in data}

FixGATCsequences <- function(df) {
  # The GATC fragments have the downside that they overlap, as they end with
  # GATC and also start with GATC. This function quickly fixes this issue.

  df.gr <- as(df, "GRanges")

  # Fix the start / end
  start(df.gr) <- start(df.gr) + 2
  end(df.gr) <- end(df.gr) - 2

  # Also trim the object to be within range
  df.gr <- trim(df.gr)

  as(df.gr, "data.frame")[, c("seqnames", "start", "end", "score")]
}

# 1) Normalized values
replicate.norm <- list()

for (s in samples) {
  
  target.norm <- list()

  for (bin in bins.with_gatc) {
    df.name <- file.path(norm_dir,
                         paste0("bin-", bin),
                         paste0(s,
                                "-",
                                bin,
                                ".norm.txt.gz"))
  
    df <- read.table(df.name,
                     sep = "\t", stringsAsFactors = FALSE,
                     col.names = c("seqnames", "start", "end", "score"))
  
    if (bin == "gatc") {
      df <- FixGATCsequences(df)
    }
    target.norm <- c(target.norm, list(df))
  }
  names(target.norm) <- bins.with_gatc
  
  replicate.norm <- c(replicate.norm, list(target.norm))
}
names(replicate.norm) <- samples


# 1b) Normalized combined values
replicate.norm.combined <- list()

for (bin in bins.with_gatc) {
  df.name <- file.path(norm_dir,
                       paste0("bin-", bin),
                       paste0(basename,
                              "-",
                              bin,
                              "-combined.norm.txt.gz"))

  df <- read.table(df.name,
                   sep = "\t", stringsAsFactors = FALSE,
                   col.names = c("seqnames", "start", "end", "score"))

  if (bin == "gatc") {
    df <- FixGATCsequences(df)
  }
  replicate.norm.combined <- c(replicate.norm.combined, list(df))
}
names(replicate.norm.combined) <- bins.with_gatc


# 2) HMM 
replicate.hmm <- list()

for (s in samples) {
  
  target.hmm <- list()

  for (bin in bins.kb) {
    df.name <- file.path(hmm_dir,
                         paste0("bin-", bin),
                         paste0(s,
                                "-",
                                bin,
                                "_HMM.txt.gz"))
  
    df <- read.table(df.name,
                     sep = "\t", stringsAsFactors = FALSE,
                     col.names = c("seqnames", "start", "end", "score"))
  
    if (bin == "gatc") {
      df <- FixGATCsequences(df)
    }
    target.hmm <- c(target.hmm, list(df))
  }
  names(target.hmm) <- bins.kb
  
  replicate.hmm <- c(replicate.hmm, list(target.hmm))
}
names(replicate.hmm) <- samples


```

Top of the normalized counts and HMM model for sample #1:

```{r read-in data printed}

head(replicate.norm[[samples[1]]][[1]])
head(replicate.hmm[[samples[1]]][[1]])

```

### Quality controls

Various quality plots will be shown below. This list is still under construction.


#### 1) Bins called

First, how many bins have reads in replicate #1 and replicate #2? 

```{r bin overlap, fig.height = 5, fig.width = 5}

for (i in 1:length(samples)) {
  print(samples[i])
  print(paste(sum(! is.na(replicate.norm[[samples[i]]][[1]][, 4])),  # NA bins
              "/",
              nrow(replicate.norm[[samples[i]]][[1]]),
              "GATC bins have reads.",
              sep = " "))
  print("")
}

```

Or the same thing in a plot

```{r bin plot, fig.height = 5, fig.width = 5}

library(Vennerable)

plotGATCOverlap <- function(replicate.norm, samples, n_bin = 1) {
  
  x <- ! is.na(replicate.norm[[samples[1]]][[n_bin]][, 4])
  y <- ! is.na(replicate.norm[[samples[2]]][[n_bin]][, 4])
    
  ab <- sum(x & y)
  a <- sum(x & !y)
  b <- sum(!x & y)
  
  if (length(samples) == 2) {
    
    w <- Venn(SetNames = samples)
    Weights(w) <- c(0, a, b, ab)
    plot(w, show = list(Faces = F))
    
  } else if (length(samples >= 3)) {
    
    z <- ! is.na(replicate.norm[[samples[3]]][[n_bin]][, 4])

    abc <- sum(x & y & z)
    ac <- sum(x & !y & z)
    bc <- sum(!x & y & z)
    c <- sum(!x & !y & z)
    
    if (length(samples) == 3) {
      
      w <- Venn(SetNames = samples)
      Weights(w) <- c(0, a, b, ab, c, ac, bc, abc)
      plot(w, show = list(Faces = F))
      
    } else if (length(samples) == 4) {
      
      q <- ! is.na(replicate.norm[[samples[4]]][[n_bin]][, 4])

      abcd <- sum(x & y & z & q)
  
      abd <- sum(x & y & !z & q)
      acd <- sum(x & !y & z & q)
      bcd <- sum(!x & y & z & q)
  
      ad <- sum(x & !y & !z & q)
      bd <- sum(!x & y & !z & q)
      cd <- sum(!x & !y & z & q)
  
      d <- sum(!x & !y & !z & q)
  
      w <- Venn(SetNames = samples)
      Weights(w) <- c(0, a, b, ab, c, ac, bc, abc, d, ad, bd, abd, cd, acd, bcd, abcd)
      plot(w, show = list(Faces = F))
      
    } else {
      stop("Sorry, this number of names is not supported yet")
    }
  }
}

for (i in 1:length(bins.with_gatc)) {
  print(paste0("Current bin size: ", bins.with_gatc[i]))
  
  tryCatch(plotGATCOverlap(replicate.norm, samples, n_bin = i),
           error = function(e) print(e))
}

```


#### 2) HMM overlap

How well does the HMM calling overlap between the replicates for different bin
sizes?

```{r HMM overlap, fig.height = 5, fig.width = 5}

library(Vennerable)

plotVennFromList <- function(df.list) {

  names.plot <- paste0("bin_", names(df.list))

  # Convert HMM list into one GRanges object
  # For now, this assumes that:
  # 1) Model go from small to large

  hmm.gr <- as(df.list[[1]],
                    "GRanges")
  names(mcols(hmm.gr)) <- names.plot[1]

  for (i in 2:length(df.list)) {

    # Determine overlap
    ovl <- findOverlaps(hmm.gr,
                        as(df.list[[i]],
                           "GRanges"),
                        type = "within")

    mcols(hmm.gr)[, names.plot[i]] <- df.list[[i]][subjectHits(ovl), "score"]

  }

  # Convert into data frame
  hmm.df <- as(hmm.gr,
               "data.frame")

  # Only select the interesting data, and only the complete cases
  df.tmp <- hmm.df[, c("seqnames", "start", "end", names.plot)]
  df.tmp <- df.tmp[complete.cases(df.tmp), ]

  # Next, create a vector of overlap
  if (length(names.plot) == 2) {

    x <- (df.tmp[, names.plot[1]] == "AD")
    y <- (df.tmp[, names.plot[2]] == "AD")

    ab <- sum(x & y)
    a <- sum(x & !y)
    b <- sum(!x & y)

    w <- Venn(SetNames = names.plot)
    Weights(w) <- c(0, a, b, ab)
    plot(w, show = list(Faces = F))

  } else if (length(names.plot) == 3) {

    x <- (df.tmp[, names.plot[1]] == "AD")
    y <- (df.tmp[, names.plot[2]] == "AD")
    z <- (df.tmp[, names.plot[3]] == "AD")

    abc <- sum(x & y & z)
    ab <- sum(x & y & !z)
    ac <- sum(x & !y & z)
    bc <- sum(!x & y & z)
    a <- sum(x & !y & !z)
    b <- sum(!x & y & !z)
    c <- sum(!x & !y & z)

    w <- Venn(SetNames = names.plot)
    Weights(w) <- c(0, a, b, ab, c, ac, bc, abc)
    plot(w, show = list(Faces = F))

  } else if (length(names.plot) == 4) {

    x <- (df.tmp[, names.plot[1]] == "AD")
    y <- (df.tmp[, names.plot[2]] == "AD")
    z <- (df.tmp[, names.plot[3]] == "AD")
    q <- (df.tmp[, names.plot[4]] == "AD")

    abcd <- sum(x & y & z & q)

    abc <- sum(x & y & z & !q)
    abd <- sum(x & y & !z & q)
    acd <- sum(x & !y & z & q)
    bcd <- sum(!x & y & z & q)

    ab <- sum(x & y & !z & !q)
    ac <- sum(x & !y & z & !q)
    ad <- sum(x & !y & !z & q)
    bc <- sum(!x & y & z & !q)
    bd <- sum(!x & y & !z & q)
    cd <- sum(!x & !y & z & q)

    a <- sum(x & !y & !z & !q)
    b <- sum(!x & y & !z & !q)
    c <- sum(!x & !y & z & !q)
    d <- sum(!x & !y & !z & q)

    w <- Venn(SetNames = names.plot)
    Weights(w) <- c(0, a, b, ab, c, ac, bc, abc, d, ad, bd, abd, cd, acd, bcd, abcd)
    plot(w, show = list(Faces = F))

  } else {
    stop("Sorry, this number of names is not supported yet")
  }

}

# Unfortunately, there is a bug in Vennerable. Try it, but don't bother if
# it doesn't work.

for (b in bins.kb) {
  print(paste0("Current bin size: ", b))
  
  hmm.list <- lapply(samples, function(x) replicate.hmm[[x]][[b]])
  
  names(hmm.list) <- samples
  
  tryCatch(plotVennFromList(hmm.list),
           error = function(e) print(e))
}


```



#### 3) Correlation between samples

First, the correlation + spearman number.

```{r correlation, fig.height = 5, fig.width = 5}

op <- par(pty="s")

n <- 10000

for (i in 1:length(bins.with_gatc)) {
  
  # Get a data frame with the observations
  df <- do.call(cbind, lapply(samples, function(x) replicate.norm[[x]][[i]][, 4]))
  df <- df[complete.cases(df), ]
  df <- data.frame(df)
  names(df) <- samples
  
  # Take a samples of this data frame
  s <- sample(1:nrow(df), min(n, nrow(df)), replace = F)
  df.reduced <- df[s, ]
  
  # Get the limits 
  limits <- c(min(df), max(df))
  
  # Plot
  my_dens <- function(data, mapping, ...) {
    ggplot(data = data, mapping=mapping) +
      geom_point(..., alpha = 0.2) +
      geom_abline(slope = 1, lty = "dashed", col = "red")
  }
  
  print(ggpairs(df.reduced,
          lower = list(continuous = my_dens)) +
    ggtitle(paste(bins.with_gatc[i],
                  "|", nrow(df.reduced), "points plotted")) +
    xlab("Normalized score (log2)") +
    ylab("Normalized score (log2)") +
    theme_bw())
  
  # plot(df, 
  #      pch = 19, col = alpha("black", 0.2),
  #      xlab = paste(samples[1], "(log 2)"),
  #      ylab = paste(samples[2], "(log 2)"),
  #      xlim = limits, ylim = limits,
  #      main = paste(bins.with_gatc[i],
  #                   "|", nrow(df), "points plotted"))
  # abline(a = 0, b = 1, lty = 2, col = "red")
  
}
par(op)


```

Also, let's plot the cross correlation. Note, this is only for replicate #1 and #2.

```{r cross correlation, fig.height = 5, fig.width = 5}

# op <- par(pty="s")
# 
# for (i in 1:length(bins.with_gatc)) {
#   x <- replicate.norm[[samples[1]]][[i]][, 4]
#   y <- replicate.norm[[samples[2]]][[i]][, 4]
#   
#   ccf(x, y, na.action = na.pass,
#       lag.max = 20,
#       main = paste(basename, 
#                    bins.with_gatc[i],
#                    sep = " - "))
# }
# 
# par(op)

```

#### 4) Autocorrelation

Multiple replicates should filter out the (technical) noise present in samples. 
If we use the autocorrelation as quality measure, can we see this?

```{r autocorrelation}

# Set-up for the ACF
acf.df <- c()

for (b in bins.with_gatc) {
  acf.vec <- c()

  for (s in samples) {
    # Where -b is the bin size and -s is the sample

    # Calculate the ACF - with the notion that NAs are simply skipped
    a <- acf(replicate.norm[[s]][[b]][, 4], lag.max = 2, na.action = na.pass, plot = F)
    acf.vec <- c(acf.vec, a$acf[2])
  }
  
  # And for the combined replicate
  a <- acf(replicate.norm.combined[[b]][, 4], lag.max = 2, na.action = na.pass, plot = F)
  acf.vec <- c(acf.vec, a$acf[2])
  
  acf.df <- cbind(acf.df, acf.vec)
}
acf.df <- data.frame(acf.df)
names(acf.df) <- bins.with_gatc
acf.df$sample <- c(samples, paste0(basename, "-combined"))


# Plot all of this
acf.df.melt <- melt(acf.df, id.vars = "sample")

ggplot(acf.df.melt, aes(x = variable, y = value, col = sample)) +
  geom_point() +
  ggtitle("Combined replicate ACF") +
  xlab("Bin size") +
  ylab("ACF") +
  scale_color_brewer(palette = "Set1") +
  theme_bw()

```

### SessionInfo

```{r}

sessionInfo()

```
